Biotic interactions structure zooplankton metacommunity dynamics following a summer heatwave

### 2023-10-13 writing/Manuscript.Rmd
revising ms based on editor feedback

\newpage
# Abstract
Despite the key role of biotic interactions in structuring ecological communities, their influence is often overlooked in predictions of how communities respond to environmental change. Here, we present an experiment that tests hypotheses based on metacommunity theory about how abiotic responses, biotic interactions, and dispersal jointly determine the response of ecological communities to environmental perturbations. We established experimental zooplankton metacommunities across spatial temperature gradients, connected by three levels of dispersal, that experienced natural temporal variation in ambient temperature. Prior to a mid-summer heatwave, community composition varied across the spatial temperature gradients. The heatwave homogenized the metacommunities and when conditions cooled, communities diverged into multiple compositional states that were not associated with temperature. These states appear to have been driven by biotic interactions that prevented the reestablishment of the pre-heatwave thermal compositional gradients. This highlights how biotic interactions can prevent metacommunities from tracking temperature changes via dispersal-facilitated species sorting.

@@

@@

We then used the Hierarchical Modelling of Species Communities (HMSC) package [@tikhonov2021] to estimate the taxon specific responses to our experimental treatments that resulted in the compositional differences that we identified using PERMANOVA. HMSC is a Bayesian generalized mixed effects model that allowed us to estimate the responses of the individual species, based on changes in abundance, to our fixed effects---temperature, dispersal, and time (sampling week)---while accounting for mesocosm and metacommunity as random effects [@ovaskainen2020]. We modelled the zooplankton community as count data using a Poisson distribution and a log-link function.

@@

We assessed the influence of our fixed and random effects using the variance partitioning method within the HMSC package [@tikhonov2021]. Because this method does not account for covariance between explanatory variables [@ovaskainen2020] it cannot properly distinguish the variance explained by temperature and dispersal from that which is explained by their interaction. Thus, we grouped temperature, dispersal, and their interaction together and estimated how much variance they explained in combination. We assessed the influence of our fixed effects on individual species based on whether they showed a positive or negative response. Significance was assessed based on whether or not the 95% credible interval overlapped zero. We also used the fitted model to estimate how species abundances should vary across a temperature range observed in the sampling weeks (19--29 \degree C), and how these response curves vary depending on the three levels of the dispersal treatment. We assessed whether species showed positive or negative co-occurrence patterns, after accounting for the fixed effects. Again we assessed significance based on whether or not the 95% credible interval overlapped zero.

@@

Evidence that biotic interactions were an important determinant of the metacommunity dynamics comes from two sources. First, several taxa exhibited negative or positive co-occurrence patterns that were not associated with our experimental treatments. This is consistent with the high compositional variation between mesocosms that was present in week 12, following the heatwave, which were not associated with the temperature treatments. It is possible that these co-occurrence patterns were the result of environmental variation that was not accounted for in the model [@blanchet2020], however non-treatment differences were minimized in our experiment and so it seems likely that these post-heatwave compositional differences were due to biotic interactions. The second source of evidence is the temporal trends in taxon abundance that were not associated with the experimental treatments. This temporal turnover in the communities is consistent with successional dynamics that are well documented in plankton communities [@sommer2012]. Our experimental design does not allow as to determine the degree to which these temporal dynamics were due to interactions between organisms that were present in the water column, or due to the emergence from resting stages in the sediment [@wisnoski2022]. Furthermore, we cannot rule out the potential that some of these dynamics may have been driven by other abiotic changes in the environment (e.g. changes in daylight). Nevertheless, regardless of the mechanism, the post-heatwave communities did not return to their pre-heatwave composition even though the taxa that were initially abundant were still present and should have been favoured by the cooler conditions. This suggests that some form of density-dependent biotic interactions were likely at play.

Our findings highlight the fact that biotic interactions are a critical component in determining how metacommunities, and communities in general respond to environmental change. This in itself should not be surprising given the central role that species play in our understanding of community ecology [@clements1916; @lotka1922; @volterra1926; @hutchinson1957]. Indeed, theory [@ives2004; @thompson2017] and experimental evidence [@davis1998; @brown2014; @alexander2015; @thompson2021] suggest that biotic interactions lead to context-dependent dynamics so that the composition of communities is rarely predictable purely based on environmental conditions. Still, most classic metacommunity theory only includes simple biotic interactions which do not result in context dependence [@loreau2003; @gonzalez2009]. Newer models that incorporate more complex interactions suggest that interactions can make it harder for species to track environmental changes through dispersal [@norberg2012; @urban2012; @thompson2017; @thompson2020]. Our experiment supports this expectation and further highlights the need to account for biotic interactions in our expectations of how species will shift their distributions in response to environmental change. This knowledge is critical for improving our ability to project how species and ecological communities will respond to future climates [@araujo2007; @svenning2014]. As our experiment demonstrates, theory to make these projections requires considering how abiotic responses, dispersal, and biotic interactions jointly govern the dynamics of ecological communities.

>>>>

\newpage
# Abstract
Despite the key role of biotic interactions in structuring ecological communities, their influence is often overlooked in predictions of how communities respond to environmental change. Here, we present an experiment that tests hypotheses based on metacommunity theory about how abiotic responses, biotic interactions, and dispersal jointly determine the response of ecological communities to environmental perturbations. We established experimental zooplankton metacommunities across spatial temperature gradients, connected by three levels of dispersal, that experienced natural temporal variation in ambient temperature. Prior to a mid-summer heatwave, community composition was predictably structured across the spatial temperature gradients and this spatial structure was strongest in metacommunities with lower rates of dispersal. The heatwave homogenized the metacommunities and when conditions cooled, communities diverged into multiple compositional states that were not associated with our experimental treatments of temperature and dispersal. Although, our experimental design did not allow us to directly assess the influence of biotic interactions, these results are consistent with the hypothesis that biotic interactions resulted in successional dynamics that prevented the reestablishment of the pre-heatwave thermal compositional gradients. These results highlight the need to better account for the role of biotic interactions in the theories that guide our expectations for how ecological communities will respond to future environmental change.

@@

While the influence of dispersal and changing abiotic conditions, including temperature, can be easily assessed through experimental manipulation [e.g., @symons2013; @limberger2019], experimentally assessing the role of biotic interactions is more challenging. In microcosms, the role of biotic interactions in shaping environmental responses has been experimentally assessed by comparing the response of species to environmental change in monoculture vs. mixture [@davis1998; @thompson2021]. Unfortunately, because each species requires its own set of replicate experimental units [e.g., 216 microcosms in @thompson2021], scaling this approach up to larger scales such as mesocosms is prohibitive. Without being able to experimentally manipulate species interactions, an alternative is to control abiotic conditions and dispersal and then use statistical models to assess if there community dynamics that cannot be explained by these factors. This includes positive or negative species co-occurrence patterns that are not related to environmental differences as well as temporal shifts in community composition that are not explained by changes in abiotic conditions [@ovaskainen2020].

@@

We then used the Hierarchical Modelling of Species Communities (HMSC) package [@tikhonov2021a] to estimate the taxon specific responses to our experimental treatments that resulted in the compositional differences that we identified using PERMANOVA. HMSC is a Bayesian generalized mixed effects model that allowed us to estimate the responses of the individual species, based on changes in abundance, to our fixed effects---temperature, dispersal, and time (sampling week)---while accounting for mesocosm and metacommunity as random effects [@ovaskainen2020]. We modelled the zooplankton community as count data using a Poisson distribution and a log-link function.

@@

We assessed the influence of our fixed and random effects using the variance partitioning method within the HMSC package [@tikhonov2021a]. Because this method does not account for covariance between explanatory variables [@ovaskainen2020] it cannot properly distinguish the variance explained by temperature and dispersal from that which is explained by their interaction. Thus, we grouped temperature, dispersal, and their interaction together and estimated how much variance they explained in combination. We assessed the influence of our fixed effects on individual species based on whether they showed a positive or negative response. Significance was assessed based on whether or not the 95% credible interval overlapped zero. We also used the fitted model to estimate how species abundances should vary across a temperature range observed in the sampling weeks (19--29 \degree C), and how these response curves vary depending on the three levels of the dispersal treatment. We assessed whether species showed positive or negative co-occurrence patterns, after accounting for the fixed effects. Again we assessed significance based on whether or not the 95% credible interval overlapped zero.

@@

Evidence that biotic interactions were an important determinant of the metacommunity dynamics comes from two sources. First, several taxa exhibited negative or positive co-occurrence patterns that were not associated with our experimental treatments. This is consistent with the high compositional variation between mesocosms that was present in week 12, following the heatwave, which were not associated with the temperature treatments. It is possible that these co-occurrence patterns were the result of environmental variation that was not accounted for in the model [@blanchet2020]. However non-treatment differences were minimized in our experiment and so it seems more likely that these post-heatwave compositional differences were due to biotic interactions. The second source of evidence is the temporal trends in taxon abundance that were not associated with the experimental treatments. This temporal turnover in the communities is consistent with successional dynamics that are well documented in plankton communities [@sommer2012]. Our experimental design does not allow us to determine the degree to which these temporal dynamics were due to interactions between organisms that were present in the water column, or due to the emergence from resting stages in the sediment [@wisnoski2022]. Furthermore, we cannot rule out the potential that some of these dynamics may have been driven by other abiotic changes in the environment (e.g. changes in daylight). Nevertheless, regardless of the mechanism, the post-heatwave communities did not return to their pre-heatwave composition even though the taxa that were initially abundant were still present and should have been favoured by the cooler conditions. This suggests that some form of density-dependent biotic interactions were likely at play.

Our findings are consistent with the hypothesis that biotic interactions are a critical component in determining how metacommunities, and communities in general respond to environmental change. This in itself should not be surprising given the central role that species play in our understanding of community ecology [@clements1916; @lotka1922; @volterra1926; @hutchinson1957]. Indeed, theory [@ives2004; @thompson2017] and experimental evidence [@davis1998; @brown2014; @alexander2015; @thompson2021] suggest that biotic interactions lead to context-dependent dynamics so that the composition of communities is rarely predictable purely based on environmental conditions. Still, most classic metacommunity theory only includes simple biotic interactions which do not result in context dependence [@loreau2003; @gonzalez2009]. Newer models that incorporate more complex interactions suggest that interactions can make it harder for species to track environmental changes through dispersal [@norberg2012; @urban2012; @thompson2017; @thompson2020]. Our experiment supports this expectation and further highlights the need to account for biotic interactions in our expectations of how species will shift their distributions in response to environmental change. This knowledge is critical for improving our ability to project how species and ecological communities will respond to future climates [@araujo2007; @svenning2014]. As our experiment demonstrates, theory to make these projections requires considering how abiotic responses, dispersal, and biotic interactions jointly govern the dynamics of ecological communities.
